---
title: "Data Wrangling 1 - HAMLA"
author: "John T McCormick (jtmccorm)"
output: html_document
---

```{r setup}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
library(tidyverse)
library(lubridate)
```

## 1. Loading

```{r}
hamla_path <- str_c(getwd(), "/data/NARA/HAMLA/", sep="")
hamla_df <- read_tsv(str_c(hamla_path, "HAMLA_1967_FS.txt", sep=""))
```

Curiously, the data-frame includes both district and hamlet evaluations. Let's take a look 

```{r}
hamla_df %>% filter(RECTP==0)
```

We can see that there are few features reliably associated with the district level observations. We'll remove those and stick to a a hamlet level analysis.

```{r}
hamla_tidy_df <- hamla_df %>% filter(RECTP==1)
```

## 2. Cleaning

### A. Identification codes

Right now there are no id codes assigned to each hamlet. Since names may be repeated in different locations, we need to fix that. This can be easily accomplished by merging the administrative id codes of higher headquarters

```{r}
hamla_tidy_df <- hamla_tidy_df %>%
  mutate(id = str_c(CHAM, PHAM, DHAM, VHAM, HHAM),
         id_country = CHAM,
         id_province = str_c(CHAM, PHAM), 
         id_district = str_c(CHAM, PHAM, DHAM),
         id_village = str_c(CHAM, PHAM, DHAM, VHAM),
         .keep = "unused") %>%
  select(id, id_country, id_province, id_district, id_village, everything())
```


### B. Date of observation

The date of each observation is stored idiosyncratically, we'll fix that as well..

```{r}
hamla_tidy_df <- hamla_tidy_df %>%
  mutate(DATE = str_c("19",str_sub(DATE, 1L, 2L), "-",str_sub(DATE, 3L, 4L), "-01"),
         date = as_date(DATE)) %>%
  select(-DATE)
```

### C. GVN and VC activities

Some of the variables contained in the date are highly idiosyncratically formatted categorical variables. A subset of these (`MLAC`, `PLAC`, `SECU`, `ADPL`) contain information on military and political activities of the VC and GVN. 

Lets tackle the VC features first and assign them descriptive names derived from the NARA documentation. Each attribute was scored between `A-E` were A represents highly favorable to the South Vietnamese and US effort and E represents highly favorable to the communist campaign. Thus, these represent more of a subjective assessment by the US advisers than raw data on the provinces.

```{r}
hamla_tidy_df <- hamla_tidy_df %>%
  mutate(
    # Extract Features from MLAC 
    vc_mil_local_forces = str_sub(MLAC, 1L, 1L), # Local Forces refer to local guerrilla/ militia forces supporting the VC
    vc_mil_external_forces = str_sub(MLAC, 2L, 2L), # External Forces refer to VC forces from outside the hamlet
    vc_mil_activities = str_sub(MLAC, 3L, 3L), # level of VC military activities in the hamlet
    vc_mil_confidence = str_sub(MLAC, 4L, 4L), # confidence the adviser has in assigned levels
    vc_mil_confidence = ifelse(vc_mil_confidence=="", NA, vc_mil_confidence),
    # Extract Features from PLAC
    vc_pol_hamlet_network = str_sub(PLAC, 1L, 1L), # the sophistication of the VC network within the hamlet
    vc_pol_village_network = str_sub(PLAC, 2L, 2L), # the sophistication of the VC network throughout the village area
    # The nework's were initial described as "infrastructure in the parlance of the Vietnam era
    vc_pol_activities = str_sub(PLAC, 3L, 3L), # level of VC political activities in the hamlet
    vc_pol_confidence = str_sub(PLAC, 4L, 4L), # confidence the adviser has in assigned levels
    vc_pol_confidence = ifelse(vc_pol_confidence=="", NA, vc_pol_confidence)) %>%
  select(-MLAC, -PLAC)
```


Now lets dissect the GVN related features. These follow a similar structure to the VC defined variables although has more specificity on the political evaluation.

```{r}
hamla_tidy_df <- hamla_tidy_df %>% 
  mutate(
    # extract features from SECU
    gvn_mil_local_forces = str_sub(SECU, 1L, 1L),
    gvn_mil_external_forces = str_sub(SECU, 2L, 2L),
    gvn_mil_activities = str_sub(SECU, 3L, 3L),
    gvn_mil_confidence = str_sub(SECU, 4L, 4L),
    gvn_mil_confidence = ifelse(gvn_mil_confidence=="", NA, gvn_mil_confidence),
    # extract features from ADPL
    gvn_pol_management = str_sub(ADPL, 1L, 1L),
    gvn_pol_popular_support = str_sub(ADPL, 2L, 2L),
    gvn_pol_info_ops = str_sub(ADPL, 3L, 3L),
    gvn_pol_confidence = str_sub(ADPL, 4L, 4L),
    gvn_pol_confidence = ifelse(gvn_pol_confidence=="", NA, gvn_pol_confidence),
  ) %>%
  select(-SECU, -ADPL)
```

### D. Lcoal Socio-Economic Factors

Columns `HEW` (Health Education, Welfare) and `ECDV` (Economic Development) have been treated in a similar method as the VC and GVN activities records above.

```{r}
hamla_tidy_df <- hamla_tidy_df %>%
  # extract features from HEW 
  mutate(
    ppl_soc_health = str_sub(HEW, 1L, 1L),
    ppl_soc_education = str_sub(HEW, 2L, 2L),
    ppl_soc_welfare = str_sub(HEW, 3L, 3L),
    ppl_soc_confidence = str_sub(HEW, 4L,4L), # advisors confidence in ratings
    ppl_soc_confidence = ifelse(ppl_soc_confidence=="", NA, ppl_soc_confidence)
  , .keep="unused") %>%  # remove old column
  # extract features from ECDV 
  mutate(
    ppl_econ_self_help = str_sub(ECDV, 1L,1L),
    ppl_econ_public_works = str_sub(ECDV, 2L, 2L),
    ppl_econ_improv_programs = str_sub(ECDV, 3L, 3L),
    ppl_econ_confidence = str_sub(ECDV, 4L, 4L), # advisors confidence in ratings
    ppl_econ_confidence = ifelse(ppl_econ_confidence=="", NA, ppl_econ_confidence)
  , .keep="unused") # remove old column
```

## E. Significant Events 

The HAMLA file contained a variable `XPROB`contains a 14 numeric string that indicates the occurence of a number of significant events within the village. 

```{r}
as.character(hamla_tidy_df$XPROB) %>% str_length()
```


### D. Misc Feature Engineering

*Remove useless features.* Several of these have no recorded meaning in the data dictionary and others are just redundant (meaning that they could be trivially calculated from other variables):

  - `+PCN`, `+SCO`, `VSZ` were created by the NIPSTRAN program (which has no referenced meaning in the dictionary)
  - `RECTP` refers to hamlet level of observation
  - `NUMB` refers to a lost "GVN identifier number"
  - `PROB` no meaning provided
  - `VCMR` no meaning provided
  - `TEAMS` no meaning provided
  - `URBAN` **a redundant indicator** of population greater than 20,000 in hamlet
  - `VALID` **redundant indicator** has the hamlet been continuously rated A through E

```{r}
hamla_tidy_df <- hamla_tidy_df %>% select(-(`+PCN`:VSZ), -VALID, -RECTP, -NUMB, -(PROB1:PROB3), -(PROB5:PROB8), -VCMR, -TEAMS)
```

Some of the features are highly idiosyncratically formatted categorical variables. These features will be removed for the time being, though future analysis would likely benefit from including these:

  - `XPROB` contains a string of indicators on 14 different types of events 
  - `MLAC`, `PLAC`, `SECU` contain information on military and political activities of the VC and GVN **(summarized in SECUR score)**
  - , `HEW`, `ECDV` contain information on the economic, administrative and health status of the local population. **(summarized in the DEVEL score)**
  
```{r}
hamla_tidy_df <- hamla_tidy_df %>% select(-(MLAC:ECDV), -XPROB)
```

As a proof of concept, let's split up a less complex idiosyncratic categorical into its components.

```{r}
hamla_tidy_df <- hamla_tidy_df %>%
  mutate(elect_num_officials = str_sub(ELECT, 1L, 2L),
         elect_reliability = str_sub(ELECT, 3L), .keep="unused")
```

Finally, lets treat non-observations as NAs...

```{r}
hamla_tidy_df <- hamla_tidy_df %>% 
  mutate(across(POPUL:CONFX, ~ if_else(.x==0, NA_real_, .x, missing=NA_real_)))
```

### D. Formatting and Organizing

```{r}
hamla_tidy_df <- hamla_tidy_df %>% 
  mutate(population = POPUL,
         score_security = SECUR,
         score_development = DEVEL,
         score_overall = CLASX,
         score_confidence = CONFX,
         hamlet = NAME, 
         village = XNAME,
         loc = POINT,
         control_Dec1966 = CNTL6,
         control_plantype = HTYPE,
         control_current = CNTL7,
         develop_status = RDSTA,
         class = CLAS,
         pop_reliability = POPRE,
         security_status = SCSTA, 
         .keep = "unused") %>%
  select(id, date, hamlet,  loc, class, village, id_village, everything())
```

Now everything looks nice and tidy!

```{r}
hamla_tidy_df
```

## 3. Geocoding (MGRS to lat-long)

The coordinates are unfortunately coded in incomplete MGRS. First, we'll standardize the format a bit and then clean to remove improperly formatted hamlet locations that can not be rectified.

```{r}
# standardize format
hamla_tidy_df <- hamla_tidy_df %>% 
  mutate(loc = paste(substr(loc,1L,2L), paste0(substr(loc,3L,5L),'0', substr(loc,6L,8L), '0')))
```


```{r}
# remove invalid locations
hamla_tidy_df <- hamla_tidy_df %>% 
  mutate(loc = ifelse(test = str_detect(loc, "^[:alpha:]{2} [:digit:]{8}$"),
                      yes = loc, no = NA_character_)) %>%
  filter(!is.na(loc))
```

We removed about a thousand observations with invalid locations. Now we can focus on adding the map group (GZD) to the coordinates.

```{r}
hamla_tidy_df <- hamla_tidy_df %>% 
  mutate(loc = case_when(
    # when vertical square identifier is in one GZD
    str_detect(substr(loc,1L,1L), "[VW]") ~ paste("48P", loc),
    str_detect(substr(loc,1L,1L), "C") ~ paste("49P", loc),
    # when vertical square identifier is split
    str_detect(substr(loc,1L,1L), "[YZXAB]") ~ case_when(
          # when  100km square is above split (Q)
          str_detect(substr(loc,1L,2L), "[YZX][DE]") ~ paste("48Q", loc),
          str_detect(substr(loc,1L,2L), "[AB]U") ~ paste("49Q", loc),
          # when 100km square is below split (P)
          str_detect(substr(loc,1L,2L), "[YZX][^C]") ~ paste("48P", loc),
          str_detect(substr(loc,1L,2L), "[AB][^T]") ~ paste("49P", loc),
          # when 100km square is split between two GZD (split by northing)
          as.numeric(substr(loc,8L,10L)) >= 707 ~ case_when( # above divider (Q row)
                # select by first grid-square code
                str_detect(substr(loc,1L,1L), "[YZX]") ~ paste("48Q", loc),
                str_detect(substr(loc,1L,1L), "[AB]") ~ paste("49Q", loc),
          ),
          as.numeric(substr(loc,8L,10L)) < 707 ~ case_when( # below divider (P row)
                # select by first grid-square code
                str_detect(substr(loc,1L,1L), "[YZX]") ~ paste("48P", loc),
                str_detect(substr(loc,1L,1L), "[AB]") ~ paste("49P", loc),
          ),
    ),
  ))
```



## 4. Export

```{r, eval=FALSE}
export_path <- str_c(getwd(), "/data/tidy/GIS_HAMLA.csv")
write_csv(hamla_tidy_df, export_path, col_names = TRUE)
```

## *Appendix:* 

### A. Known Data Inconsistencies:
 
 1. Location (MGRS)
    * not full MGRS, no map group provided
    * not standard across the hamlets record
    * only six-digit accuracy (but not all are six-digit accuracy)
    
2. Hamlet & Village Name
    * inconsistency between dash `-` and space ` `
    * 
    


